---
title: "Final Report"
author: "Diana Ballesteros (deg2163), Chris Crowe (clc2229), Tanvi Jain (uni: tj2383) "
date: "December 4th, 2018"
---

#Analysis of a Longitudinal Study on Aging Conducted in Ireland 

Motivation: This project aims to look at the effect of sociodemographic indicators on mental health among a cohort of Irish residents aged 50+ years old.

Related Work: This project is a broad examination of the implications of aging on physical and mental health. The inspiration for this analysis was drawn from the group members research interests including depression, aging, and chronic diseases. The following link to a page from the American Psychological Association was used as a reference for the variables we chose to analyze. https://www.apa.org/helpcenter/aging-depression.aspx 

Questions: 
Our main questions - Is physical health associated with mental health? Does loneliness change over time among widowed men? What is the proportion of various ICD-10 diseases among this population?

Evolution of questions: As we created the graphs we decided to also explore variations according to sex (male and female) and different marital statuses (widowed, married, divorced, cohabitating, separated, and single). For some graphs, we also explored the change over time by stratifying according to wave.

Evolution of methods: In our initial proposal we wanted to look at `Quality of Life`. However, the nature of this variable seemed ambiguous to us so instead we looked at variables such as `mental health`, `physical health`, and `loneliness` as a proxy for quality of life. We also planned to use a hexagonal heat map for graph 1 but according to the distribution of our data this map was not visually aesthetic, instead we used a geom_count to display the concentrations of data within a scatterplot.

Loading the necessary libraries:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(plotly)
library(viridis)
library(plotly)
library(tidyverse)
library(MASS)

select = dplyr::select
```

Instructions for importing the data: download the data from Google Drive to your local R environment (this requires you to sign into your Google account) then reference the data from a local path.
```{r load_data}
load("./data/Wave 1/DS0001/34315-0001-Data.rda")
load("./data/Wave 2/DS0001/37105-0001-Data.rda")
load("./data/Wave 3/DS0001/37106-0001-Data.rda")

wave_1_data = da34315.0001
wave_2_data = da37105.0001
wave_3_data = da37106.0001

nrow(wave_1_data)
nrow(wave_2_data)
nrow(wave_3_data)
```


Tidying the data: 
```{r}
## Create functions to clean variables
clean_wave = function(x) {
  
  x %>% 
    mutate(
      wave = case_when(
        str_detect(wave, "_1$") ~ 1,
        str_detect(wave, "_2$") ~ 2,
        str_detect(wave, "_3$") ~ 3
    )
  )
  
}
clean_physical = function(x) {
  
  x %>% 
    mutate(
      physical = case_when(
        physical_value == "(1) Excellent" ~ "Excellent", 
        physical_value == "(2) Very good" ~ "Very Good",
        physical_value == "(3) Good" ~ "Good", 
        physical_value == "(4) Fair" ~ "Fair", 
        physical_value == "(5) Poor" ~ "Poor",
        physical_value == "(01) Excellent" ~ "Excellent", 
        physical_value == "(02) Very Good" ~ "Very Good",
        physical_value == "(03) Good" ~ "Good", 
        physical_value == "(04) Fair" ~ "Fair", 
        physical_value == "(05) Poor" ~ "Poor"),
      physical = factor(physical, levels = c("Poor", "Fair", "Good", "Very Good", "Excellent"))
    )
  
}
clean_mental = function(x) {
  
  x %>% 
    mutate(
      mental = case_when(
        mental_value == "(1) Excellent" ~ "Excellent", 
        mental_value == "(2) Very good" ~ "Very Good",
        mental_value == "(3) Good" ~ "Good", 
        mental_value == "(4) Fair" ~ "Fair", 
        mental_value == "(5) Poor" ~ "Poor",
        mental_value == "(01) Excellent" ~ "Excellent", 
        mental_value == "(02) Very Good" ~ "Very Good",
        mental_value == "(03) Good" ~ "Good", 
        mental_value == "(04) Fair" ~ "Fair", 
        mental_value == "(05) Poor" ~ "Poor"),
      mental = factor(mental, levels = c("Poor", "Fair", "Good", "Very Good", "Excellent"))
    )
}
clean_marital = function(x) {
  
  x %>% 
    mutate(
      marital = case_when(
        CS006 == "(1) Married" ~ "Married",
        CS006 == "(2) Living with a partner as if married" ~ "Living w/ Partner",
        CS006 == "(3) Single (never married)" ~ "Never married",
        CS006 == "(4) Separated" ~ "Separated",
        CS006 == "(5) Divorced" ~ "Divorced",
        CS006 == "(6) Widowed" ~ "Widowed"),
      marital = factor(marital, levels = c("Married", "Living w/ Partner", "Never married",
                                           "Separated", "Divorced", "Widowed"))
    )
  
}
clean_sex = function(x) {
  
   x %>% 
    mutate(
      sex = case_when(
        SEX == "(1) Male" ~ "Male",
        SEX == "(2) Female" ~ "Female")
    )
  
}

## Assign data from each wave to a new object where common variable names are given unique names
wave_1_data = da34315.0001 %>% 
  mutate(MHUCLA_LONELINESS_1 = MHUCLA_LONELINESS,
         PH001_1 = PH001,
         PH002_1 = PH002)
 
wave_2_data = da37105.0001 %>% 
  mutate(MHUCLA_LONELINESS_2 = MHUCLA_LONELINESS,
         PH001_2 = PH001,
         PH002_2 = PH002
  )
      
wave_3_data = da37106.0001 %>% 
  mutate(MHUCLA_LONELINESS_3 = MHUCLA_LONELINESS,
         PH001_3 = PH001,
         PH002_3 = PH002
  )
## Data for Physical Health vs. Mental Health
## Tidy data for self-report physical health
overall_1 = wave_1_data %>%  
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(PH001_1, PH001_2, PH001_3, SEX.x, CS006, ID) %>% 
  gather(wave, physical_value, PH001_1:PH001_3) %>% 
  clean_wave()
  
## Tidy data for self-report mental health
overall_2 = wave_1_data %>%  
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(PH002_1, PH002_2, PH002_3, SEX.x, CS006, ID) %>% 
  gather(wave, mental_value, PH002_1:PH002_3) %>% 
  clean_wave()
## Tidy data for loneliness
overall_3 = wave_1_data %>%  
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3, SEX.x, CS006, ID) %>% 
  gather(wave, loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  clean_wave()
## Tidy data for ICD codes
overall_4 = wave_1_data %>%  
  merge(wave_3_data, by = "ID") %>% 
  select(ICD10_01:ICD10_16, ID, SEX.x, CS006) %>% 
  mutate(SEX = SEX.x) %>% 
  clean_sex() %>% 
  clean_marital() %>% 
  select(-SEX.x, -SEX, -CS006)
  
## Combine data for self-report physical and mental health, apply functions to create consistent, clean labels for variables
final = overall_1 %>% 
  merge(overall_2, by = c("ID", "wave")) %>% 
  merge(overall_3, by = c("ID", "wave")) %>% 
  select(ID, wave, SEX.x, CS006.x, physical_value, mental_value, loneliness_value) %>% 
  clean_physical() %>% 
  clean_mental() %>% 
  mutate(
    CS006 = CS006.x,
    SEX = SEX.x
  ) %>% 
  clean_marital() %>% 
  clean_sex() %>% 
  select(-physical_value, -mental_value, -SEX, -CS006, -SEX.x, -CS006.x)

## Prepare data for loneliness plot
plot_3 = 
  wave_1_data %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(SEX.x, CS006, ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3)  %>% 
  mutate(
    SEX = SEX.x
  ) %>% 
  clean_sex() %>% 
  clean_marital() %>% 
  select(-SEX.x, -CS006, -SEX)
```

Data Source: The data was obtained from The Irish Longitudinal Study on Ageing (TILDA) Waves 1, 2, and 3. Here is a link to the dataset https://www.icpsr.umich.edu/icpsrweb/ICPSR/series/726 . We did not delete any participants from our dataset to ensure completeness of our code, in case a participant who is null right now responds in a future wave. Our sample size for wave 1 was n =`r nrow(wave_1_data)`, for wave 2 n = `r nrow(wave_2_data)`, for wave 3 n = `r nrow(wave_3_data)`.

Exploratory Analysis - creating graphs

Graph 1:
```{r countplot}
countplot = wave_3_data %>% 
  select(PH001, PH002, SEX) %>% 
  mutate(physical = case_when(
    PH001 == "(01) Excellent" ~ "Excellent", 
    PH001 == "(02) Very Good" ~ "Very Good",
    PH001 == "(03) Good" ~ "Good", 
    PH001 == "(04) Fair" ~ "Fair", 
    PH001 == "(05) Poor" ~ "Poor"), 
    
  mental = case_when(
    PH002 == "(01) Excellent" ~ "Excellent", 
    PH002 == "(02) Very Good" ~ "Very Good",
    PH002 == "(03) Good" ~ "Good", 
    PH002 == "(04) Fair" ~ "Fair", 
    PH002 == "(05) Poor" ~ "Poor"),
  
  sex = case_when(
    SEX == "(1) Male" ~ "Male",
    SEX == "(2) Female" ~ "Female"
  ),
  
  physical = factor(physical, levels = c("Excellent", "Very Good", "Good", "Fair", "Poor")), 
  
  mental = factor(mental, levels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
        ) %>% 
  select(-PH001, -PH002, -SEX)

plot_1 = countplot %>%
  ggplot(aes(x = physical, y = mental, color = ..n..)) +
  geom_count(alpha = 0.8) +
  labs(
    x = "Self-Rated Physical Health",
    y = "Self-Rated Mental Health"
  ) +
  facet_grid(~sex) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))


```

For the first plot, we cleaned the values under the variables of interest using the `case_when` function and used the `factor` function to level categorical ordinal variables, physical health and mental health, in decreasing order. We chose not to delete null values from our dataset in case participants chose to respond in future waves. We then created a scatter plot showing the concentration of responses according to dot size. We used `shiny` to stratify this plot by sex, marital status, and across waves to observe trends over time. 

We ran a chi-squared test between mental health and physical health. First, we checked to make sure the expected cell counts were greater than 5. Then we ran code to calculate the chi-squared test statistic in order to determine if there is an association between mental health and physical heatlh.
```{r}
final =
  final %>%
  filter(wave == 3)
 
chi = chisq.test(table(final$mental, final$physical))
 
observed = chi$observed %>%
  knitr::kable()
 
expected = chi$expected %>%
  knitr::kable()
 
results = chi %>%
  broom::tidy() %>%
  janitor::clean_names() %>%
  mutate(df = parameter,
         chi_stat = statistic) %>%
  select(chi_stat, df, p_value) %>%
  knitr::kable()

results
```

At the 5% level of significance we have sufficient evidence to conclude that there is an association between mental health and physical health (X^2 = , p-value = ).

Graph 2:
```{r spaghetti}
wave_1_data = wave_1_data %>% 
  mutate(MHUCLA_LONELINESS_1 = MHUCLA_LONELINESS)

wave_2_data = wave_2_data %>% 
  mutate(MHUCLA_LONELINESS_2 = MHUCLA_LONELINESS)

wave_3_data = wave_3_data %>% 
  mutate(MHUCLA_LONELINESS_3 = MHUCLA_LONELINESS)

spaghetti = 
  merge(wave_1_data, wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 

  select(ID, SEX.x, CS006, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  arrange(ID) %>% 
  mutate(sex = case_when(
    SEX.x == "(1) Male" ~ "Male",
    SEX.x == "(2) Female" ~ "Female"
  ),
    marital_status = case_when(
      CS006 == "(1) Married" ~ "Married",
      CS006 == "(2) Living with a partner as if married" ~ "Cohabitation",
      CS006 == "(3) Single(never married)" ~ "Single",
      CS006 == "(4) Separated" ~ "Separated",
      CS006 == "(5) Divorced" ~ "Divorced",
      CS006 == "(6) Widowed" ~ "Widowed"
)) %>% 
  filter(marital_status == "Widowed") %>% 
  select(-SEX.x, -CS006) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
      wave == "MHUCLA_LONELINESS_1" ~ 1, 
      wave == "MHUCLA_LONELINESS_2" ~ 2, 
      wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  janitor::clean_names()

spaghetti %>% 
  filter(sex == "Male", loneliness_value != "NA") %>% 
  ggplot(aes(x = wave, y = loneliness_value, color = id)) +
  geom_line() +
  theme_bw() +
  theme(legend.position = "none")
```

(DELETING THIS GRAPH 2?)

Graph 3:
```{r bar_graph}
# ICD-10: bar graph of proportions of each ICD-10 code to determine which ICD-10 diagnosis accounts for the greatest burden of disease in our population. We can potentially facet by sex to see if the burden is different for males and females. 
bar_graph = 
  wave_3_data %>% 
  select(ID, ICD10_01:ICD10_16) %>% 
  summarize(total_01 = sum(ICD10_01), total_02 = sum(ICD10_02), total_03 = sum(ICD10_03), 
            total_04 = sum(ICD10_04), total_05 = sum(ICD10_05), total_06 = sum(ICD10_06), 
            total_07 = sum(ICD10_07), total_08 = sum(ICD10_08), total_09 = sum(ICD10_09),
            total_10 = sum(ICD10_10), total_11 = sum(ICD10_11), total_12 = sum(ICD10_12),
            total_13 = sum(ICD10_13), total_14 = sum(ICD10_14), total_15 = sum(ICD10_15),
            total_16 = sum(ICD10_16)) %>% 
  gather(key = icd_code, value = total, total_01:total_16) %>% 
  separate(icd_code, into = c("total_char","icd_code"), sep = "_") %>% 
  select(-total_char) %>% 
  mutate(burden = total/sum(total)*100) %>% 
  mutate(disease = case_when(
    icd_code == "01" ~ "Infectious diseases",
    icd_code == "02" ~ "Neoplasms",
    icd_code == "03" ~ "Blood diseases",
    icd_code == "04" ~ "Nutritional/metabolic diseases",
    icd_code == "05" ~ "Mental/behavioral disorders",
    icd_code == "06" ~ "Nervous system diseases",
    icd_code == "07" ~ "Eye diseases",
    icd_code == "08" ~ "Ear diseases",
    icd_code == "09" ~ "Circulatory system diseases",
    icd_code == "10" ~ "Respiratory system diseases",
    icd_code == "11" ~ "Digestive system diseases",
    icd_code == "12" ~ "Skin diseases",
    icd_code == "13" ~ "Musculoskeletal system diseases",
    icd_code == "14" ~ "Genitourinary system diseases",
    icd_code == "15" ~ "Perinatal conditions",
    icd_code == "16" ~ "Congenital malformations"
  ))
 

plot_2 = bar_graph %>% 
  ggplot(aes(x = reorder(disease, -burden), y = burden)) +
  geom_bar(stat = "identity") + 
  labs(
    title = "Proportion of people with specified diseases",
    x = "Disease",
    y = "Percentage"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  ) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 7))

```

In this plot we calculated the sum of people who had each ICD-10 disease in wave 3 in order to calculate the proportion of people with each disease, allowing us to visualize the burden of each disease in this population. Again, we used `case_when` to clean the values under the variables of interest. We then created a bar graph to show the proportions and ordered the bars according to decreasing burden. We used `shiny` to further stratify by sex and marital status.

Graph 2 (real):
```{r}
### Baseline 0

baseline_loneliness_0 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 0 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_0 = baseline_loneliness_0 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 1

baseline_loneliness_1 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 1 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_1 = baseline_loneliness_1 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 2

baseline_loneliness_2 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 2 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_2 = baseline_loneliness_2 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 3


baseline_loneliness_3 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 3 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_3 = baseline_loneliness_3 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 4

baseline_loneliness_4 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 4 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_4 = baseline_loneliness_4 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)


### Baseline 5

baseline_loneliness_5 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 5 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_5 = baseline_loneliness_5 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 6

baseline_loneliness_6 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 6 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_6 = baseline_loneliness_6 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 7

baseline_loneliness_7 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 7 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_7 = baseline_loneliness_7 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 8

baseline_loneliness_8 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 8 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_8 = baseline_loneliness_8 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 9

baseline_loneliness_9 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 9 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_9 = baseline_loneliness_9 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 10

baseline_loneliness_10 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 10 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_10 = baseline_loneliness_10 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE) + ylim(0, 10)

plot_3 = (baseline_0 + baseline_1 + baseline_2 + baseline_3 + baseline_4 + baseline_5 + baseline_6 + baseline_7 + baseline_8 + baseline_9)
```

For the second plot we renamed the loneliness variable so that it was consistent across the three waves and then merged the three waves of data by ID. Again, we used `case_when` to clean the values under the variables of interest. We then created a spaghetti plot to show the change in loneliness across waves. A large change me made once we saw this chaotic plot was that we decided to depict the different trends across waves for each baseline loneliness value (scale from 0-10). Although there is no one at loneliness 10 we kept this in our code for future replication of this code, if a participant were to respond as 10. We created separate datasets for each loneliness value at baseline and then at each wave thereafter. The resulting spaghetti plot shows the trend across waves in across loneliness values. The use of `shiny` allowed us to further stratify by sex and marital status. We also incorporated a color grid to depict the different trends.



Chi-Square Test: 
```{r}
final =
  final %>%
  filter(wave == 3)
 
chi = chisq.test(table(final$mental, final$physical))
 
observed = chi$observed %>%
  knitr::kable()
 
expected = chi$expected %>%
  knitr::kable()
 
results = chi %>%
  broom::tidy() %>%
  janitor::clean_names() %>%
  mutate(df = parameter,
         chi_stat = statistic) %>%
  select(chi_stat, df, p_value) %>%
  knitr::kable()
 
## observed
## expected
## results
```


Discussion:
Graph 1: We anticipated that physical health and mental health would be correlated and from graph 1 we discovered that this was true. Those who felt they had good physical health also had good mental health and vice versa. We also found a larger concentration of older adults reporting `Very Good` or `Good` physical and mental health. This trend remained consistent across sex and over time. Unfortunately, this means that there was minimal improvement in physical and mental health, but also minimal decline in physical and mental health.

Chi-Squared test: Since our graph depicted a correlation between mental health and physical health we anticipated an association. Our null hypothesis was that there is no association between mental health and physical health and the laternate hypothesis was that there is an association between mental health and physical health. Setting alpha to 0.05 and calculating the chi-squared test statistic and p-value, which is less than 0.05, we determined that there is enough evidence to reject the null hypothesis and conclude that there is an association between mental health and physical health, which is reflected in graph 1.

Graph 3: Our hypothesis for this plot was more broad in that we expected the proportion of people with chronic diseases to be higher, simply because the cohort is older and in Ireland, which may not have a high prevalence of infectious, or perinatal diseases and conditions. The bar graph specified that the 4 diseases with the highest proportion were circulatory system diseases, eye diseases, nutritional/metabolic diseases, and musculoskeletal system diseases. As predicted there were no incident cases of infectious diseases or perinatal conditions. Additionally, there were no cases of congenital malformations, ear diseases, or skin conditions. Interestingly, across sex, we noticed that the most prevalent disease among men was circulatory system diseases but among women it was nutritional/metabolic diseases.

Graph 2 (real): We expected that loneliness would increase over time for widowed men more than widowed women. According to graph 2 we found this to be true. For widowed men, those with a loneliness of 2 or higher seemed to decrease their loneliness across waves and those with a loneliness of 1 seemed to increase their loneliness across waves. As we expected, widowed women stayed at relatively the same loneliness level across all waves with the exception of an an outlier at level 8 loneliness. In contrast, married men seemed to decrease in loneliness at every baseline point, while women stayed relatively the same from wave 1 to wave 3.

Regression Model and results: We wanted to conduct a simple linear regression testing the association between disease type and proportion of people with burden. 
