---
title: "Final Report"
---

#Analysis of a Longitudinal Study on Aging Conducted in Ireland 

Motivation: This project aims to look at the effect of sociodemographic indicators on mental health among a cohort of Irish residents aged 50+ yrs.

Related Work: This project is a broad examination of the implications of aging on physical and mental health. The inspiration for this analysis was drawn from the group members researhc interests including depression, aging, and chronic diseases. The following link to a page from the American Psychological Association was used as a reference for the variables we chose to analyze. https://www.apa.org/helpcenter/aging-depression.aspx 

Questions: 
Our main questions: Is physical health associated with mental health? Does loneliness change over time among widowed men?

Evolution of questions: As we created the graphs we decided to also explore variations according to sex. For example, we also explored loneliness among widowed women over time and saw that loneliness increased more for men compared to women.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(patchwork)
library(plotly)
```

Data Source: The data was obtained from The Irish Longitudinal Study on Ageing (TILDA) Waves 1, 2, and 3. Here is a link to the dataset https://www.icpsr.umich.edu/icpsrweb/ICPSR/series/726

Importing data

```{r load_data}
load("./data/Wave 1/DS0001/34315-0001-Data.rda")
load("./data/Wave 2/DS0001/37105-0001-Data.rda")
load("./data/Wave 3/DS0001/37106-0001-Data.rda")

wave_1_data = da34315.0001
wave_2_data = da37105.0001
wave_3_data = da37106.0001

```

Exploratory Analysis:

Creating Graphs

```{r countplot}
countplot = wave_3_data %>% 
  select(PH001, PH002, SEX) %>% 
  mutate(physical = case_when(
    PH001 == "(01) Excellent" ~ "Excellent", 
    PH001 == "(02) Very Good" ~ "Very Good",
    PH001 == "(03) Good" ~ "Good", 
    PH001 == "(04) Fair" ~ "Fair", 
    PH001 == "(05) Poor" ~ "Poor"), 
    
  mental = case_when(
    PH002 == "(01) Excellent" ~ "Excellent", 
    PH002 == "(02) Very Good" ~ "Very Good",
    PH002 == "(03) Good" ~ "Good", 
    PH002 == "(04) Fair" ~ "Fair", 
    PH002 == "(05) Poor" ~ "Poor"),
  
  sex = case_when(
    SEX == "(1) Male" ~ "Male",
    SEX == "(2) Female" ~ "Female"
  ),
  
  physical = factor(physical, levels = c("Excellent", "Very Good", "Good", "Fair", "Poor")), 
  
  mental = factor(mental, levels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
        ) %>% 
  select(-PH001, -PH002, -SEX)

plot_1 = countplot %>%
  ggplot(aes(x = physical, y = mental, color = ..n..)) +
  geom_count(alpha = 0.8) +
  labs(
    x = "Self-Rated Physical Health",
    y = "Self-Rated Mental Health"
  ) +
  facet_grid(~sex) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 40, hjust = 1))

ggplotly(plot_1)

```

```{r spaghetti}
wave_1_data = wave_1_data %>% 
  mutate(MHUCLA_LONELINESS_1 = MHUCLA_LONELINESS)

wave_2_data = wave_2_data %>% 
  mutate(MHUCLA_LONELINESS_2 = MHUCLA_LONELINESS)

wave_3_data = wave_3_data %>% 
  mutate(MHUCLA_LONELINESS_3 = MHUCLA_LONELINESS)

spaghetti = 
  merge(wave_1_data, wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 

  select(ID, SEX.x, CS006, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  arrange(ID) %>% 
  mutate(sex = case_when(
    SEX.x == "(1) Male" ~ "Male",
    SEX.x == "(2) Female" ~ "Female"
  ),
    marital_status = case_when(
      CS006 == "(1) Married" ~ "Married",
      CS006 == "(2) Living with a partner as if married" ~ "Cohabitation",
      CS006 == "(3) Single(never married)" ~ "Single",
      CS006 == "(4) Separated" ~ "Separated",
      CS006 == "(5) Divorced" ~ "Divorced",
      CS006 == "(6) Widowed" ~ "Widowed"
)) %>% 
  filter(marital_status == "Widowed") %>% 
  select(-SEX.x, -CS006) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
      wave == "MHUCLA_LONELINESS_1" ~ 1, 
      wave == "MHUCLA_LONELINESS_2" ~ 2, 
      wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  janitor::clean_names()

spaghetti %>% 
  filter(sex == "Male", loneliness_value != "NA") %>% 
  ggplot(aes(x = wave, y = loneliness_value, color = id)) +
  geom_line() +
  theme_bw() +
  theme(legend.position = "none")
```

```{r bar_graph}
# ICD-10: bar graph of proportions of each ICD-10 code to determine which ICD-10 diagnosis accounts for the greatest burden of disease in our population. We can potentially facet by sex to see if the burden is different for males and females. 
bar_graph = 
  wave_3_data %>% 
  select(ID, ICD10_01:ICD10_16) %>% 
  summarize(total_01 = sum(ICD10_01), total_02 = sum(ICD10_02), total_03 = sum(ICD10_03), 
            total_04 = sum(ICD10_04), total_05 = sum(ICD10_05), total_06 = sum(ICD10_06), 
            total_07 = sum(ICD10_07), total_08 = sum(ICD10_08), total_09 = sum(ICD10_09),
            total_10 = sum(ICD10_10), total_11 = sum(ICD10_11), total_12 = sum(ICD10_12),
            total_13 = sum(ICD10_13), total_14 = sum(ICD10_14), total_15 = sum(ICD10_15),
            total_16 = sum(ICD10_16)) %>% 
  gather(key = icd_code, value = total, total_01:total_16) %>% 
  separate(icd_code, into = c("total_char","icd_code"), sep = "_") %>% 
  select(-total_char) %>% 
  mutate(burden = total/sum(total)*100) %>% 
  mutate(disease = case_when(
    icd_code == "01" ~ "Infectious diseases",
    icd_code == "02" ~ "Neoplasms",
    icd_code == "03" ~ "Blood diseases",
    icd_code == "04" ~ "Nutritional/metabolic diseases",
    icd_code == "05" ~ "Mental/behavioral disorders",
    icd_code == "06" ~ "Nervous system diseases",
    icd_code == "07" ~ "Eye diseases",
    icd_code == "08" ~ "Ear diseases",
    icd_code == "09" ~ "Circulatory system diseases",
    icd_code == "10" ~ "Respiratory system diseases",
    icd_code == "11" ~ "Digestive system diseases",
    icd_code == "12" ~ "Skin diseases",
    icd_code == "13" ~ "Musculoskeletal system diseases",
    icd_code == "14" ~ "Genitourinary system diseases",
    icd_code == "15" ~ "Perinatal conditions",
    icd_code == "16" ~ "Congenital malformations"
  ))
 

plot_2 = bar_graph %>% 
  ggplot(aes(x = reorder(disease, -burden), y = burden)) +
  geom_bar(stat = "identity") + 
  labs(
    title = "Proportion of people with specified diseases",
    x = "Disease",
    y = "Proportion (%)"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  ) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 7))
  
ggplotly(plot_2)

```

```{r}
### Baseline 0

baseline_loneliness_0 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 0 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_0 = baseline_loneliness_0 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 1

baseline_loneliness_1 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 1 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_1 = baseline_loneliness_1 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 2

baseline_loneliness_2 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 2 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_2 = baseline_loneliness_2 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 3


baseline_loneliness_3 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 3 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_3 = baseline_loneliness_3 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 4

baseline_loneliness_4 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 4 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_4 = baseline_loneliness_4 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)


### Baseline 5

baseline_loneliness_5 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 5 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_5 = baseline_loneliness_5 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 6

baseline_loneliness_6 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 6 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_6 = baseline_loneliness_6 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 7

baseline_loneliness_7 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 7 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_7 = baseline_loneliness_7 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 8

baseline_loneliness_8 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 8 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_8 = baseline_loneliness_8 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 9

baseline_loneliness_9 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 9 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_9 = baseline_loneliness_9 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE)

### Baseline 10

baseline_loneliness_10 = wave_1_data %>% 
  filter(MHUCLA_LONELINESS_1 == 10 & SEX == "(1) Male" & CS006 == "(6) Widowed") %>% 
  select(ID, SEX, CS006, MHUCLA_LONELINESS_1) %>% 
  merge(wave_2_data, by = "ID") %>% 
  merge(wave_3_data, by = "ID") %>% 
  select(ID, MHUCLA_LONELINESS_1, MHUCLA_LONELINESS_2, MHUCLA_LONELINESS_3) %>% 
  gather(key = wave, value = loneliness_value, MHUCLA_LONELINESS_1:MHUCLA_LONELINESS_3) %>% 
  mutate(wave = case_when(
    wave == "MHUCLA_LONELINESS_1" ~ 1,
    wave == "MHUCLA_LONELINESS_2" ~ 2,
    wave == "MHUCLA_LONELINESS_3" ~ 3
  )) %>% 
  filter(loneliness_value != "NA")

baseline_10 = baseline_loneliness_10 %>% 
  ggplot(aes(x = wave, y = loneliness_value)) +
    geom_smooth(se = FALSE) + ylim(0, 10)

plot_3 = (baseline_0 + baseline_1 + baseline_2 + baseline_3 + baseline_4 + baseline_5 + baseline_6 + baseline_7 + baseline_8 + baseline_9)

ggplotly(plot_3)
```

Discussion:

Graph 1: We anticipated that those who felt they had good physical health would also have good mental health and vice versa. From Graph 1
Self-rated physical and mental health are correlated
Most older adults in this population report having good or very good physical and mental health
Same for males and females
Graph 2:
Of all the incident diseases, the top 4 are:
Circulatory system diseases
Eye diseases
Nutritional/metabolic diseases
Musculoskeletal system diseases
There were no incident cases of:
Congenital malformations
Ear diseases
Infectious diseases
Perinatal conditions
Skin conditions
Graph 3:


